---
title: "Manual Curation Evaluation"
date: '`r Sys.Date()`'
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(ggpubr)
```

# Objective
Develop figure summarizing evaluation / manual curation results for draft benchmark

# Approach
Using googlesheet summary table, input data table cleaned first in `tidying_data.Rmd` 

_Note from Slack_ RE: spreadsheet source 
    @nolson for the evaluation results figure, I think you'll want to use my curation results in columns T-U, unless they are blank, in which case you should use columns Q-S, in the GRCh37andGRCh38 tab in https://docs.google.com/spreadsheets/d/1Pn7WP78JfWKCO2Df31n_4gzDOwtP69flgDyyjeBS6JE/edit?usp=sharing.  This'll be a bit more complicated than previous figs, since we have the CommonFP/FN/FP_FN rows, so happy to chat about this when you get to it (edited) 


# Loading Data
```{r}

## Callset method info
callset_meta_df <- read_tsv(here("data", "benchmark_evaluation", "dataset_names.tsv"))

## Manual Curation Results
bench_eval_df <- read_tsv(here("data","tidy","manual_curation.tsv"))

## Annotating with Tech
anno_bench_eval_df <- bench_eval_df %>% 
    left_join(callset_meta_df)
    
no_common_df <-  anno_bench_eval_df %>%
    filter(callset != "Common")

common_df <- anno_bench_eval_df %>%
    filter(callset == "Common") %>% 
    mutate(var_type = case_when(str_detect(truth_gt, "SNP") ~ "SNP",
                            str_detect(truth_gt, "INDEL") ~ "INDEL",
                            str_detect(query_gt, "SNP") ~ "SNP",
                            str_detect(query_gt, "INDEL") ~ "INDEL",
                            nchar(ref_base) == 1 & nchar(alt_base) == 1 ~ "SNP",
                            nchar(ref_base) > 1 | nchar(alt_base) > 1 ~ "INDEL",
                            TRUE ~ "DEBUGME")
    )

mc_df <- no_common_df #bind_rows(no_common_df, common_df)
```
__TODO:__ Work out how to incorporate COMMON variants

Code from V4 manuscript
https://gitlab.nist.gov/gitlab/jmw7/v4_manuscript/-/blob/master/scripts/manual_curation.R 


# Figure generation
```{r}
# Panel A #####################################################################
panel_a_df <- mc_df %>% 
        mutate(curation = paste(benchmark_correct, query_correct)) %>% 
        group_by(callset_id, var_type, fp_or_fn) %>% 
        mutate(total_vars = n(),
               fp_or_fn = if_else(fp_or_fn == "FP_FN", "genotype", fp_or_fn)) %>% 
        group_by(callset_id, var_type, fp_or_fn, curation, total_vars) %>%
        summarise(curation_count = n()) %>% 
        mutate(curation_prop = curation_count/ total_vars) %>% 
        filter(curation == "yes no") %>%
        ungroup() %>% 
        mutate(callset_id = fct_reorder(callset_id,curation_prop, min)) %>% 
        mutate(var_type = factor(var_type, levels = c("SNP","INDEL")))

panel_a <- ggplot(panel_a_df) +
        geom_bar(aes(x = callset_id, y = curation_prop, fill = fp_or_fn),
                 color = "grey40",
                 size = 0.25,
                 stat = "identity",
                 position = "dodge") +
        geom_hline(aes(yintercept = 0.5), linetype = 2) +
        facet_wrap(~var_type, ncol = 3) +
        scale_fill_brewer(palette = "Paired",type = "qual") + 
        theme_bw() +
        labs(x = "Callset",
             y = "Proportion",
             fill = "Discrepancy Type") +
        theme(legend.position = "bottom") + 
        coord_flip()

## Panel B #####################################################################
grey_vars_df <- mc_df %>%
        mutate(curation = paste(benchmark_correct, query_correct)) %>%
        filter(curation != "yes no") %>%
        group_by(benchmark_correct) %>%
        mutate(count = n()) %>%
        ungroup() %>%
        mutate(benchmark_correct = fct_reorder(benchmark_correct,
                                               count, .desc = TRUE)) 

panel_b <- ggplot(grey_vars_df) +
        geom_bar(aes(x = benchmark_correct, 
                     fill = query_correct),
                 color = "grey40",
                 size = 0.25) +
        scale_fill_brewer(palette = 2, type = "qual") + 
        theme_bw() +
        labs(x = "Benchmark Curation", y = "Variants", 
             fill = "Query Curation") + 
        theme( legend.position = "bottom")

## Panel C #####################################################################
unsure_vars <- mc_df %>%
        mutate(curation = paste(benchmark_correct, query_correct)) %>%
        filter(benchmark_correct == "unsure") %>% 
        group_by(callset_id) %>%
        mutate(count = n()) %>%
        ungroup() %>%
        mutate(callset_id = fct_reorder(callset_id,count))  %>%
        mutate(query_correct = factor(query_correct, 
                                      levels = c("no", "unsure","yes"))
        )

panel_c <-  ggplot(unsure_vars) +
        geom_bar(aes(x = callset_id, fill = query_correct),
                 color = "grey40", size = 0.25) + 
        coord_flip() + 
        theme_bw() + 
        scale_fill_manual(values = c(no = "#1b9e77", unsure = "#7570b3")) +
        labs(y = "Variants", x = "Callset", fill = "Query Curation") +
        theme(legend.position = "bottom")

## Combining Panels ############################################################
full_fig <- ggpubr::ggarrange(panel_a,  
                              ggarrange(panel_b, panel_c, 
                                        ncol = 2, 
                                        legend = "bottom",
                                        labels = c("B","C"),
                                        common.legend = TRUE),
                              nrow = 2,
                              labels = "A", heights = c(4,3))

```

```{r}
full_fig
```


## Saving to file
```{r}
ggsave(filename = here("figures", "figS9_manual_curation.png"), 
       full_fig, width = 6, height = 6)

ggsave(filename = here("figures", "figS9_manual_curation.pdf"), 
       full_fig, width = 6, height = 6)

```


# Session Information
## System Information
```{r}
sessioninfo::platform_info()
```


## Package Versions
```{r}
sessioninfo::package_info() %>% 
    filter(attached = TRUE) %>% 
    select(package, loadedversion, date, source) %>%
    knitr::kable(booktabs = TRUE, row.names = FALSE)
```