---
title: "Supplemental Gene Metrics Table"
date: '`r Sys.Date()`'
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

# Objective
Generate summary table with gene stratification overlap and benchmark coverage information for genes in the CMRG benchmark and genes excluded from the benchmark due to incomplete assemblies.

Table description
- ref
- gene
- coords
- start_coordinates_plus_segdups_plus_flanking
- end_coordinates_plus_segdups_plus_flanking
- in_mrg
- v4_lt90
- mandelker
- cosmic
- high_priority
- gene+overlapping_segdup+flank coverage by the dip.bed
    - bp_hifiasm_dip.bed
    - fraction_hifiasm_dip.bed
- gene+overlapping_segdup+flank coverage stratifications
    - fraction_in_ccs_ont_elliptical_outlier_cnv_stratification
    - fraction_in_assembly_cnv_stratification
    - fraction_in_segmental_duplications_stratification
    - fraction_in_Tier1andTier2_SV_0.6_slop150_stratification
    - fraction_in_reference_gaps_stratification

# Approach
- Getting gene coordinates from full gene bed
- Bind with coords + flank
- Row match verification flank includes full gene coords
- annotate with coverage and other information

# Loading Input Data Files
Gene Coordinates
```{r}
gene_coords_df <- read_tsv(here("data", "tidy", "asm_coverage.tsv.gz"))
```


# Annotations
## V4 Coverage and MRG inclusion information
```{r}
v4_included_df <- here("data", "tidy", "bench_coverage.tsv.gz") %>% 
    read_tsv(col_types = "ccccccddl") %>% 
     filter(region == "gene", benchmark == "v4") %>% 
    select(-benchmark, -bench_type, -region) %>% 
    rename(v4_gene_bp = bp_cov, v4_gene_frac = cov)

glimpse(v4_included_df)
```

## Assembly Coverage
```{r}
asm_gene_cov_df <- read_tsv(here("data","tidy","asm_coverage.tsv.gz"),
                            col_types = "ccciiciiid"
                            ) %>% 
    select(-giab_id) %>%
    filter(region == "gene-plus-flank") %>%
    rename(start_flank = start, end_flank = end)
```

## Strat Overlap
```{r}
strat_overlap_df <- here("data", "tidy", "strat_overlap.tsv.gz") %>%
    read_tsv() %>% 
    filter(region == "gene-plus-flank") %>% 
    rename(start_flank = start, end_flank = end) %>% 
    select(ref, chrom, start_flank, end_flank, strat, overlap) %>% 
    distinct() %>% 
    mutate(strat = str_replace_all(strat, "-","_")) %>% 
    pivot_wider(names_from = strat, names_prefix = "cov_", values_from = overlap)
```

```{r}
glimpse(strat_overlap_df)
```

# Combined Table

Table description
X- ref
X- gene
X- coords
X- start_coordinates_plus_segdups_plus_flanking
X- end_coordinates_plus_segdups_plus_flanking
X- in_mrg
X- v4_lt90
- mandelker
- cosmic
- high_priority
X- gene+overlapping_segdup+flank coverage by the dip.bed
X    - bp_hifiasm_dip.bed
X    - fraction_hifiasm_dip.bed
- gene+overlapping_segdup+flank coverage stratifications
    - fraction_in_ccs_ont_elliptical_outlier_cnv_stratification
    - fraction_in_assembly_cnv_stratification
    - fraction_in_segmental_duplications_stratification
    - fraction_in_Tier1andTier2_SV_0.6_slop150_stratification
    - fraction_in_reference_gaps_stratification


```{r}
bench_inclusion_df <- gene_coords_df %>% 
    left_join(v4_included_df)%>% 
    mutate(v4_lt90 = v4_gene_frac <= 0.9)
```

```{r}
glimpse(bench_inclusion_df)
```

```{r}
gene_stats_tbl <- bench_inclusion_df %>% 
    left_join(asm_gene_cov_df) %>% 
    select(-region, -region_size) %>% 
    rename(asm_cov_bp = bp_cov, 
           asm_cov_frac = overlap)
```

```{r}
glimpse(gene_stats_tbl)
```

```{r}
gene_stats_tbl <- left_join(gene_stats_tbl, strat_overlap_df) %>% 
    select(-chrom)
```
```{r}
glimpse(gene_stats_tbl)
```

Supperset gene list
```{r}
challenging_genes <- bench_inclusion_df %>% 
    filter(v4_lt90 == TRUE,
           str_detect(coords, "X|Y|MT", negate = TRUE)) %>% 
    .$gene %>% unique()

length(challenging_genes)
```


```{r}
challenging_genes_df <- gene_stats_tbl %>% 
    filter(gene %in% challenging_genes)

glimpse(challenging_genes_df)    
```
```{r}
write_tsv(challenging_genes_df, file = here("data","challenging_genes_tbl.tsv.gz"))
```

```{r}
write_tsv(gene_stats_tbl, file = here("data","gene_stat_tbl.tsv.gz"))
```

# Session Information
## System Information
```{r}
sessioninfo::platform_info()
```


## Package Versions
```{r}
sessioninfo::package_info() %>% 
    filter(attached = TRUE) %>% 
    select(package, loadedversion, date, source) %>%
    knitr::kable(booktabs = TRUE, row.names = FALSE)
```